An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-04-14 cal_demo/ms.tex
Breaking up lines in cal_demo paper for easier tracking of changes.

\section{Introduction}
In order to satisfy the survey speeds required for precision cosmology as well as searches for fast radio transients, radio astronomy is undergoing a paradigm shift toward interferometers consisting of hundreds to thousands of small, widefield antennas. Many arrays with this design are already built or under construction including the Hydrogen Epoch of Reionization Array\footnote{http://reionization.org} (HERA), the Murchison Widefield Array (MWA; \citealt{tin13,bow13}), the Donald C. Backer Precision Array for Probing the Epoch of Reionization (PAPER; \citealt{par10}), the LOw Frequency ARray (LOFAR; \citealt{van13}), the Canadian Hydrogen Intensity Mapping Experiment (CHIME,\citealt{ban14}), the Long Wavelength Array (LWA, \citealt{ell13}), and the low frequency Square Kilometer Array (SKA1-Low \citealt{mel13}).

@@

A handful of prototype direct imaging correlators have been tested on arrays including the Basic Element for SKA Training II (BEST-2) array \citep{fos14}, the Omniscope \citep{zhe14}, and an earlier pulsar timing experiment at GHz frequencies \citep{oto94, dai00}. Each of these are examples of so-called FFT correlators -- a subclass of direct imaging correlators which rely on identical antennas with restricted placement, which allows the FFT to be performed without gridding. We recently released the E-field Parallel Imaging Correlator \citep[EPIC;][]{thy15c}, which is a software implementation of the Modular Optimal Frequency Fourier \citep[MOFF;][]{mor11} imaging algorithm. This architecture leverages the software holography/A-transpose framework to grid electric field data streams before performing the spatial FFT, allowing for an optimal map without placing constraints on array layout or requiring identical antennas \citep{mor09,bha08,teg97a}.

@@

Here we present the E-field Parallel Imaging Calibration (EPICal) algorithm -- a novel solution to the calibration problem, which can be integrated into direct imaging correlators and scales only as the number of antennas, $\mathcal{O}(\Nant)$. This method uses a correlation of the uncalibrated antenna signal stream with an output image pixel from the backend of the correlator to solve for the complex gains of the antennas. Because the calibration must be applied before gridding and imaging, our solution requires an iterative approach where the data from one time series is used to update the gains which are applied to the following time series. An example implementation of the algorithm is available with the EPIC software package\footnote{http://github.com/nithyanandan/EPIC}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig1.pdf}
\caption{Data flow for the MOFF algorithm, reproduced from \citealt{thy15c}. The time ordered electric fields are collected by the antenna elements then frequency transformed in the F-engine. The calibration is applied, then the electric field measurements are gridded using the antenna voltage patterns. The imager performs a spatial 2D FFT on the gridded data before squaring and averaging. The accumulated images are then written to disk.}
\label{fig:moff_flow}
\end{center}
\end{figure}

@@

We show schematically the process of calibrating a direct imaging correlator in figure~\ref{fig:schematic}. Computationally expensive steps that must be performed ``on-chip" are shown inside the gray box. The uncalibrated antenna signals are tapped out after the F-engine and correlated against the output image pixel of interest (i.e., equation~\ref{eq:Cna} is ``on-chip''). The correlated values are then passed off-chip to estimate the gains using equation~\ref{eq:cal_solution}, and additional fitting if desired. The gains are then passed back to the correlator to update the calibration for subsequent integration intervals.

@@

For our unknown gains, we create a set of random complex numbers where the amplitude is drawn from approximately gaussian distribution centered around 1 with width 0.25, and completely random phase. These are our ``true gains", and we apply them to the frequency-domain simulated antenna electric fields as in equation~\ref{eq:apply_gain}. Our analysis is blind to these values until the end of the process to check accuracy. The gain estimates are initialized with unity, $g^{(0)}_a=1$.

@@

The estimated gain amplitudes for the simulations are shown in figure~\ref{fig:sim_amp}. The quantity plotted is the magnitude of the estimated gains over the true gains, $\left|g^{(n)}_a/g^T_a\right|$, which places all antennas on the same scale. We can see the amplitudes converge toward their true values around the same time as the phases (iteration $\sim$15). At the beginning of calibration we can see the value of the damping factor. At $n=0$, a couple of gains are shown to have abnormally high amplitude estimates, notably one about 3.3 times its true value (red line). These unbalanced high estimates caused the entire set of gains to be under estimated at $n=1$, even with a damping factor of 0.35. By $n=5$ the unbalanced amplitudes have been damped out and the calibration continues. Without the damping factor, the oscillation seen in the first couple iterations would have been significantly larger and taken much longer to fade out.

@@

Finally, we found that seven antennas\footnote{LWA antenna IDs 48, 85, 124, 148, 203, 217, and 244 were flagged.} produced unstable gain solutions, and in fact corrupted the entire array. We therefore flagged these antennas in our analysis, resulting in a total of 248 antennas to calibrate. In each calibration loop, we form $C_{i,\hat{\boldsymbol{s}}_0}, i=1,2,\ldots N_a$ and update our gain estimates over 10 timestamps (51.2 ms). We iterate the loop 30 times for a total of 1.536 seconds. The results of this calibration experiment are shown in figures~\ref{fig:data_phase}~--~\ref{fig:data_images}.

@@

To simulate receiver noise, we add a gaussian distributed complex random number to each antenna electric field measurement at each time stamp according to equation~\ref{eq:apply_gain}. The level of the receiver noise, $\sigma_r = \left<\left|\widetilde{n}_a(f,t)\right|^2\right>$, is varied in different simulation runs. We include two limiting cases, where the receiver noise power is subdominant to the sky power ($\sigma_r = 10.0$~Jy), and where the receiver dominates the noise ($\sigma_r=100.0$~Jy).

@@

For the EPICal calibration loops, we initialize our gain estimates with the true values (${g^{(0)}_a=g^T_a=1}$), and allow the estimate to be corrupted by the noise through ten iterations of the calibration loop. We adopt a damping factor $\damp=0.35$. Because EPICal updates the gain estimate at each calibration loop, \tcal, but retains a memory of previous iterations through the damping factor, the total integration time is not straightforward. We define an effective integration time by considering the relative weights of each previous $\Cna$ contributing to the current $g^{(n)}_a$ estimate, each with integration time \tcal. In the limit $n\rightarrow \infty$, the geometric series converges to an effective integration time of
\begin{equation}
\teff = \tcal \times \frac{1+\damp}{1-\damp}.
\end{equation}
With the damping factor we adopted here, $\teff \approx 2.1 \times \tcal$.

We simultaneously form simulated visibilities by correlating all pairs of antenna E-field measurements (including the receiver noise). The E-fields are correlated for a duration equal to the effective integration time of the EPICal loop for comparison. We then find the visibility-based gain estimates by minimizing
\begin{equation}\label{eq:vis_cal}
\chi^2 = \sum_a\sum_{b\ne a} \left|\V_{ab}-g_a g_b^* \V'_{ab}\right|^2,
\end{equation}
using both versions of the model visibilities described above.

@@

\textbf{Improving sky model.} In this work we used a perfect sky model (for simulation), or a very simple sky model (for LWA data). In principle the images produced by the EPIC correlator can be used to improve the sky model used in calibration -- similar to a major loop in self-calibration. This can be especially useful for compact, widefield arrays which require a model of both compact and diffuse sources over a large patch of sky. In addition to requiring the sky model, the calibration also requires this model to be attenuated by the primary beam, which can be difficult to measure at the precision necessary \citep[e.g.][]{neb15,vir14,thy15b}. However, the direct imaging correlator provides exactly the model necessary in real time, and can be iterated over to improve the images and gain solutions.

>>>>

\section{Introduction}
In order to satisfy the survey speeds required for precision cosmology as well as searches for 
fast radio transients, radio astronomy is undergoing a paradigm shift toward interferometers 
consisting of hundreds to thousands of small, widefield antennas. Many arrays with this design 
are already built or under construction including the Hydrogen Epoch of Reionization Array
\footnote{http://reionization.org} (HERA), the Murchison Widefield Array (MWA; 
\citealt{tin13,bow13}), the Donald C. Backer Precision Array for Probing the Epoch of 
Reionization (PAPER; \citealt{par10}), the LOw Frequency ARray (LOFAR; \citealt{van13}), the 
Canadian Hydrogen Intensity Mapping Experiment (CHIME,\citealt{ban14}), the Long 
Wavelength Array (LWA, \citealt{ell13}), and the low frequency Square Kilometer Array (SKA1-
Low \citealt{mel13}).

@@

A handful of prototype direct imaging correlators have been tested on arrays including the 
Basic Element for SKA Training II (BEST-2) array \citep{fos14}, the Omniscope \citep{zhe14}, 
and an earlier pulsar timing experiment at GHz frequencies \citep{oto94, dai00}. Each of these 
are examples of so-called FFT correlators -- a subclass of direct imaging correlators which rely 
on identical antennas with restricted placement, which allows the FFT to be performed without 
gridding. We recently released the E-field Parallel Imaging Correlator \citep[EPIC;][]{thy15c}, 
which is a software implementation of the Modular Optimal Frequency Fourier \citep[MOFF;][]
{mor11} imaging algorithm. This architecture leverages the software holography/A-transpose 
framework to grid electric field data streams before performing the spatial FFT, allowing for an 
optimal map without placing constraints on array layout or requiring identical antennas 
\citep{mor09,bha08,teg97a}.

@@

Here we present the E-field Parallel Imaging Calibration (EPICal) algorithm -- a novel solution 
to the calibration problem, which can be integrated into direct imaging correlators and scales 
only as the number of antennas, $\mathcal{O}(\Nant)$. This method uses a correlation of the 
uncalibrated antenna signal stream with an output image pixel from the backend of the 
correlator to solve for the complex gains of the antennas. Because the calibration must be 
applied before gridding and imaging, our solution requires an iterative approach where the data 
from one time series is used to update the gains which are applied to the following time series. 
An example implementation of the algorithm is available with the EPIC software package
\footnote{http://github.com/nithyanandan/EPIC}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig1.pdf}
\caption{Data flow for the MOFF algorithm, reproduced from \citealt{thy15c}. The time ordered 
electric fields are collected by the antenna elements then frequency transformed in the F-
engine. The calibration is applied, then the electric field measurements are gridded using the 
antenna voltage patterns. The imager performs a spatial 2D FFT on the gridded data before 
squaring and averaging. The accumulated images are then written to disk.}
\label{fig:moff_flow}
\end{center}
\end{figure}

@@

We show schematically the process of calibrating a direct imaging correlator in 
figure~\ref{fig:schematic}. Computationally expensive steps that must be performed ``on-chip" 
are shown inside the gray box. The uncalibrated antenna signals are tapped out after the F-
engine and correlated against the output image pixel of interest (i.e., equation~\ref{eq:Cna} is 
``on-chip''). The correlated values are then passed off-chip to estimate the gains using 
equation~\ref{eq:cal_solution}, and additional fitting if desired. The gains are then passed back 
to the correlator to update the calibration for subsequent integration intervals.

@@

For our unknown gains, we create a set of random complex numbers where the amplitude is 
drawn from approximately gaussian distribution centered around 1 with width 0.25, and 
completely random phase. These are our ``true gains", and we apply them to the frequency-
domain simulated antenna electric fields as in equation~\ref{eq:apply_gain}. Our analysis is 
blind to these values until the end of the process to check accuracy. The gain estimates are 
initialized with unity, $g^{(0)}_a=1$.

@@

The estimated gain amplitudes for the simulations are shown in figure~\ref{fig:sim_amp}. The 
quantity plotted is the magnitude of the estimated gains over the true gains, $\left|g^{(n)}_a/
g^T_a\right|$, which places all antennas on the same scale. We can see the amplitudes 
converge toward their true values around the same time as the phases (iteration $\sim$15). At 
the beginning of calibration we can see the value of the damping factor. At $n=0$, a couple of 
gains are shown to have abnormally high amplitude estimates, notably one about 3.3 times its 
true value (red line). These unbalanced high estimates caused the entire set of gains to be 
under estimated at $n=1$, even with a damping factor of 0.35. By $n=5$ the unbalanced 
amplitudes have been damped out and the calibration continues. Without the damping factor, 
the oscillation seen in the first couple iterations would have been significantly larger and taken 
much longer to fade out.

@@

Finally, we found that seven antennas\footnote{LWA antenna IDs 48, 85, 124, 148, 203, 217, 
and 244 were flagged.} produced unstable gain solutions, and in fact corrupted the entire array. 
We therefore flagged these antennas in our analysis, resulting in a total of 248 antennas to 
calibrate. In each calibration loop, we form $C_{i,\hat{\boldsymbol{s}}_0}, i=1,2,\ldots N_a$ and 
update our gain estimates over 10 timestamps (51.2 ms). We iterate the loop 30 times for a 
total of 1.536 seconds. The results of this calibration experiment are shown in figures~
\ref{fig:data_phase}~--~\ref{fig:data_images}.

@@

To simulate receiver noise, we add a gaussian distributed complex random number to each 
antenna electric field measurement at each time stamp according to 
equation~\ref{eq:apply_gain}. The level of the receiver noise, $\sigma_r = \left<\left|
\widetilde{n}_a(f,t)\right|^2\right>$, is varied in different simulation runs. We include two limiting 
cases, where the receiver noise power is subdominant to the sky power ($\sigma_r = 
10.0$~Jy), and where the receiver dominates the noise ($\sigma_r=100.0$~Jy).

@@

For the EPICal calibration loops, we initialize our gain estimates with the true values 
(${g^{(0)}_a=g^T_a=1}$), and allow the estimate to be corrupted by the noise through ten 
iterations of the calibration loop. We adopt a damping factor $\damp=0.35$. Because EPICal 
updates the gain estimate at each calibration loop, \tcal, but retains a memory of previous 
iterations through the damping factor, the total integration time is not straightforward. We define 
an effective integration time by considering the relative weights of each previous $\Cna$ 
contributing to the current $g^{(n)}_a$ estimate, each with integration time \tcal. In the limit $n
\rightarrow \infty$, the geometric series converges to an effective integration time of
\begin{equation}
\teff = \tcal \times \frac{1+\damp}{1-\damp}.
\end{equation}
With the damping factor we adopted here, $\teff \approx 2.1 \times \tcal$.

We simultaneously form simulated visibilities by correlating all pairs of antenna E-field 
measurements (including the receiver noise). The E-fields are correlated for a duration equal to 
the effective integration time of the EPICal loop for comparison. We then find the visibility-
based gain estimates by minimizing
\begin{equation}\label{eq:vis_cal}
\chi^2 = \sum_a\sum_{b\ne a} \left|\V_{ab}-g_a g_b^* \V'_{ab}\right|^2,
\end{equation}
using both versions of the model visibilities described above.

@@

\textbf{Improving sky model.} In this work we used a perfect sky model (for simulation), or a 
very simple sky model (for LWA data). In principle the images produced by the EPIC correlator 
can be used to improve the sky model used in calibration -- similar to a major loop in self-
calibration. This can be especially useful for compact, widefield arrays which require a model of 
both compact and diffuse sources over a large patch of sky. In addition to requiring the sky 
model, the calibration also requires this model to be attenuated by the primary beam, which 
can be difficult to measure at the precision necessary \citep[e.g.][]{neb15,vir14,thy15b}. 
However, the direct imaging correlator provides exactly the model necessary in real time, and 
can be iterated over to improve the images and gain solutions.
